---
title: "Fitting Exercise"
---

# Load and summarize data
```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(GGally)
library(tidymodels)

mavo <- read_csv("Mavoglurant.csv")

glimpse(mavo)
dim(mavo)
```

# Convert data to plot
```{r}
ggplot(mavo, aes(x = TIME, y = DV, group = ID, color = factor(DOSE))) +
  geom_line(alpha = 0.6) +
  labs(color = "Dose") +
  theme_minimal()

#only keeping observations with OCC = 1
mavo_occ1 <- mavo %>%
  filter(OCC == 1)

table(mavo_occ1$OCC)
```

# Creates a data frame with time that is not equal to 0 and computes the sum of DV
```{r}
mavo_no0 <- mavo_occ1 %>%
  filter(TIME != 0)

Y_df <- mavo_no0 %>%
  group_by(ID) %>%
  summarize(Y = sum(DV, na.rm = TRUE))
```

# Creates a data frame with time = 0
```{r}
baseline_df <- mavo_occ1 %>%
  filter(TIME == 0)
```

# Joins the tow data frames together
```{r}
final_df <- baseline_df %>%
  left_join(Y_df, by = "ID")

dim(final_df)
```

# Coverts race and sex to variable factors and cleans data
```{r}
clean_df <- final_df %>%
  mutate(
    SEX = factor(SEX),
    RACE = factor(RACE)
  ) %>%
  select(Y, DOSE, AGE, SEX, RACE, WT, HT)

saveRDS(clean_df, file = "clean-data.rds")

summary(clean_df)
```

# Summary of data
```{r}
clean_df %>%
  summarize(
    n = n(),
    mean_Y = mean(Y),
    sd_Y = sd(Y),
    mean_age = mean(AGE),
    mean_WT = mean(WT),
    mean_HT = mean(HT)
  )
```
This summary table shows that there are 120 indivduals and it shows the the mean and standard deviation of drug exposure, and the mean od age, WT, and HT.

# Summary grouped by dose 
```{r}
  clean_df %>%
  group_by(DOSE) %>%
  summarize(
    n = n(),
    mean_Y = mean(Y),
    sd_Y = sd(Y),
    mean_WT = mean(WT),
    mean_AGE = mean(AGE)
  )
```
This summary table shows the relationship between dose and exposure. As dose increases, total exposure decreases.

# Summary grouped by sex
```{r}
  clean_df %>%
  group_by(SEX) %>%
  summarize(
    n = n(),
    mean_Y = mean(Y),
    mean_WT = mean(WT)
  )
```
This summary table shows the data grouped by sex. Sex 1 has 104 individuals and sex 2 has 16 individuals. 

# Plot y versus dose
```{r}
ggplot(clean_df, aes(x = factor(DOSE), y = Y)) +
  geom_boxplot() +
  labs(x = "Dose", y = "Total Drug (Y)") +
  theme_minimal()
```
This boxplot shows total drug exposure versus dose; the higher the dose is associated with drug exposure.

# Plot y versus age
```{r}
  ggplot(clean_df, aes(x = AGE, y = Y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()
```
This scatterplot shows total drug exposure versus age; there is no relationship between the two variables.

# Distribution of Y
```{r}
ggplot(clean_df, aes(x = Y)) +
  geom_histogram(bins = 20) +
  theme_minimal()
```
This histogram shows total drug exposure skewed to the right. This means that most individuals have exposure values between 1500 and 3500, but a few individuals have high exposure values.

# Distribution of HT
```{r}
  ggplot(clean_df, aes(x = HT)) +
  geom_histogram(bins = 15) +
  theme_minimal()
```
The histogram shoes height with a normal distribution with most idividuals within 1.60 to 1.90 meters.

# Correlation matrix 
```{r}
numeric_df <- clean_df %>%
  select(Y, AGE, WT, HT)

cor(numeric_df)
```
This correlation matrix shows the strength and whether the relationship is positive or negative between total drug exposure, age, height, and weight. +1 menas there is a positive relationship, 0 means there is no relationship, and -1 means there is a negative relationship.

# Correlation matrix plot plot
```{r}
ggpairs(clean_df,
        columns = c("Y", "AGE", "WT", "HT"),
        aes(color = factor(DOSE)))
```
This correlation matrix plot shows the same relationships as the correlation matrix table.

# Model fitting
```{r}
set.seed(123)

data_split <- initial_split(clean_df, prop = 0.8, strata = SEX)

train_data <- training(data_split)
test_data  <- testing(data_split)
```

# Fitting the linear model to the continous outcome of Y using dose as the main predictor
```{r}
lm_spec <- linear_reg() %>%
  set_engine("lm")

lm_dose <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE)

lm_dose_fit <- fit(lm_dose, data = train_data)

pred_dose <- predict(lm_dose_fit, test_data) %>%
  bind_cols(test_data)

  metrics_dose <- pred_dose %>%
  metrics(truth = Y, estimate = .pred)

metrics_dose
```

# Fitting the linear model to the continous outcome of Y using all the predictors
```{r}
lm_all <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE + AGE + SEX + RACE + WT + HT)

lm_all_fit <- fit(lm_all, data = train_data)

pred_all <- predict(lm_all_fit, test_data) %>%
  bind_cols(test_data)

metrics_all <- pred_all %>%
  metrics(truth = Y, estimate = .pred)

metrics_all
```

# Fitting the logistic model to the continous outcome of sex using dose as the main predictor
```{r}
log_spec <- logistic_reg() %>%
  set_engine("glm")

log_dose <- workflow() %>%
  add_model(log_spec) %>%
  add_formula(SEX ~ DOSE)

log_dose_fit <- fit(log_dose, data = train_data)

pred_log_dose <- predict(log_dose_fit, test_data, type = "prob") %>%
  bind_cols(
    predict(log_dose_fit, test_data),
    test_data
  )

pred_log_dose %>%
  metrics(truth = SEX, estimate = .pred_class)

pred_log_dose %>%
  roc_auc(truth = SEX, .pred_1)
```

# Fitting the logistic model to the continous outcome of Y using all the predictors
```{r}
log_all <- workflow() %>%
  add_model(log_spec) %>%
  add_formula(SEX ~ DOSE + AGE + RACE + WT + HT)

log_all_fit <- fit(log_all, data = train_data)

pred_log_all <- predict(log_all_fit, test_data, type = "prob") %>%
  bind_cols(
    predict(log_all_fit, test_data),
    test_data
  )

pred_log_all %>%
  metrics(truth = SEX, estimate = .pred_class)

pred_log_all %>%
  roc_auc(truth = SEX, .pred_1)
```

The first linear model is using dose as the only predictor of total drug exposure. It shows that the higher the dose, the higher the drug exposure is. 

The second linear model is using all variables (age, dose, sex, race, height, and weight) as predictors of total drug exposure. It showed that dose was the strongest predictor of total drug exposure. 

The first logistic model is using dose as the only predictor of sex. This model produced low ROC-AUC and accuracy values because there is no relationship between dose and sex.

The second logistic model is using all the variables as predictors of sex. This model has higher ROC-AUC and accuracy values compared to the first model becuase height and weight can potentially be predictors of sex.

# Model Improvement 
Excluding race as a variable
```{r}
rngseed = 1234

clean_df <- final_df %>%
  mutate(
    SEX = factor(SEX)
  ) %>%
  select(Y, DOSE, AGE, SEX, WT, HT)

dim(clean_df)
summary(clean_df)
```

Setting test and train data
```{r}
set.seed(rngseed)

data_split <- initial_split(clean_df, prop = 0.75)

train_data <- training(data_split)
test_data  <- testing(data_split)

nrow(train_data)
nrow(test_data)
```

# Linear models
```{r}
lm_spec <- linear_reg() %>%
  set_engine("lm")
```

# Linear Model with Dose
```{r}
lm_dose_workflow <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE)

lm_dose_fit <- fit(lm_dose_workflow, data = train_data)

lm_dose_pred <- predict(lm_dose_fit, new_data = train_data) %>%
  bind_cols(train_data %>% select(Y))

metrics(lm_dose_pred, truth = Y, estimate = .pred)
```

# Linear Model with all Predictors
```{r}
lm_all_workflow <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE + AGE + SEX + WT + HT)

lm_all_fit <- fit(lm_all_workflow, data = train_data)

lm_all_pred <- predict(lm_all_fit, new_data = train_data) %>%
  bind_cols(train_data %>% select(Y))

metrics(lm_all_pred, truth = Y, estimate = .pred)
```

# Null model 
```{r}
mean_y <- mean(train_data$Y)

null_pred <- train_data %>%
  mutate(.pred = mean_y)

rmse_null <- null_pred %>%
  rmse(truth = Y, estimate = .pred)

rmse_null
```

The null model, predicts the mean outcome for all observations, had the highest RMSE being 948. This RMSE value indicates that the predictive preformace of this model is poor. The model that used dose as the only predictor had a lower RMSE value (703), this suggests that dose explains variability in the outcome. The model that uses all predictors had the lowest RMSE (627), which indicates the best preformance. 

```{r}
set.seed(rngseed)

# creates 10 fold CV split
folds <- vfold_cv(train_data, v = 10)
```

# Model using Dose
```{r}
lm_dose_workflow <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE)
```

# Model Using All Predictors 
```{r}
lm_all_workflow <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE + AGE + SEX + WT + HT)
```

# Cross Validation
```{r}
  dose_cv <- fit_resamples(
  lm_dose_workflow,
  resamples = folds,
  metrics = metric_set(rmse)
)

all_cv <- fit_resamples(
  lm_all_workflow,
  resamples = folds,
  metrics = metric_set(rmse)
)
```

# RMSE and Standard Validation 
```{r}
dose_cv_results <- collect_metrics(dose_cv)
all_cv_results  <- collect_metrics(all_cv)

dose_cv_results
all_cv_results
```

The RMSE values from the cros validation are a little higher than those from the training data, this could be because of less overfitting. This gives us a more realistic idea of the models preformance on new data. The model that used all the predictors still had the lowest RMSE value, followed by the model that used dose as the only predictor, and the null model had the highest RMSE value. The standard error values suggest that there is some variability in how the model preforms across folds, but the differences are consistent. 

# Different Value for the Random Seed
```{r}
set.seed(999)

folds2 <- vfold_cv(train_data, v = 10)

dose_cv2 <- fit_resamples(lm_dose_workflow, resamples = folds2, metrics = metric_set(rmse))
all_cv2  <- fit_resamples(lm_all_workflow, resamples = folds2, metrics = metric_set(rmse))

collect_metrics(dose_cv2)
collect_metrics(all_cv2)
```

Changing the random seed resulted in slightly different RMSE values, this could be due to variation in how the data was split into the folds. The pattern stayed the same with the model that used all the predictors having the lowest RMSE value, and the null model having the highest RMSE value. 

# This section added by Riley Herber

## Predictions for each model
```{r}
# Dose model predictions
dose_pred <- predict(lm_dose_fit, new_data = train_data) %>%
  bind_cols(train_data %>% select(Y)) %>%
  mutate(model = "Dose Model")

# All predictor model predictions
all_pred <- predict(lm_all_fit, new_data = train_data) %>%
  bind_cols(train_data %>% select(Y)) %>%
  mutate(model = "All Predictors Model")

# Null model predictions
null_pred <- train_data %>%
  mutate(.pred = mean(train_data$Y),
         model = "Null Model") %>%
  select(Y, .pred, model)
```

## Combine predictions into one dataframe
```{r}
predictions_df <- bind_rows(
  dose_pred %>% select(Y, .pred, model),
  all_pred %>% select(Y, .pred, model),
  null_pred
)
```

## Observed vs Predicted

```{r}
ggplot(predictions_df, aes(x = Y, y = .pred, color = model, shape = model)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(0, 5000), ylim = c(0, 5000)) +
  labs(
    x = "Observed Values",
    y = "Predicted Values",
    color = "Model",
    shape = "Model"
  ) +
  theme_minimal()
```

For the model that includes only dose as a predictor, the predicted values fall along three horizontal lines because the DOSE variable only takes three distinct values in the dataset. 

## Plot All Predictors Model

```{r}
all_residuals <- predict(lm_all_fit, new_data = train_data) %>%
  bind_cols(train_data %>% select(Y)) %>%
  mutate(residual = .pred - Y)

ggplot(all_residuals, aes(x = .pred, y = residual)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_cartesian(ylim = c(-2000, 2000)) +
  labs(
    x = "Predicted Values",
    y = "Residuals (Predicted - Observed)",
    title = "Residuals vs Predicted Values for Full Model"
  ) +
  theme_minimal()
```

# Model predictions and uncertainty

```{r}
library(rsample)
library(purrr)

set.seed(rngseed) 

# 100 bootstrap resamples
dat_bs <- bootstraps(train_data, times = 100)
```

## Fit the model on each bootstrap and make predictions for the original training data
```{r}
# store predictions
pred_list <- map(dat_bs$splits, function(split) {

dat_sample <- rsample::analysis(split)
  
  # fit full model on bootstrap sample
  fit_bs <- lm(Y ~ DOSE + AGE + SEX + WT + HT, data = dat_sample)
  
  # predict on original training data
  predict(fit_bs, newdata = train_data)
})
```

## Combine predictions into a matrix
```{r}
pred_bs <- do.call(rbind, pred_list)
dim(pred_bs) 
```

## Compute median and 89% confidence intervals
```{r}
preds_ci <- apply(pred_bs, 2, quantile, probs = c(0.055, 0.5, 0.945)) %>%
  t() %>%
  as.data.frame()

colnames(preds_ci) <- c("lower", "median", "upper")
preds_ci$observed <- train_data$Y
preds_ci$predicted <- predict(lm_all_fit, new_data = train_data)$.pred
```

## Plot observed vs. predicted with uncertainty
```{r}
library(ggplot2)

ggplot(preds_ci, aes(x = observed)) +
  geom_point(aes(y = predicted), color = "black", alpha = 0.7) +        # original point estimate
  geom_point(aes(y = median), color = "blue", alpha = 0.5) +             # median from bootstrap
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0, color = "red", alpha = 0.3) +  # CI
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(0, 5000), ylim = c(0, 5000)) +
  labs(
    x = "Observed Values",
    y = "Predicted Values",
    title = "Observed vs Predicted Values with 89% Bootstrap Confidence Intervals"
  ) +
  theme_minimal()
```

The plot of observed versus predicted values with 89% bootstrap confidence intervals shows that the full linear model predicts total drug exposure reasonably well. The black points represent the original model predictions, and most lie close to the 45° line, indicating general agreement between observed and predicted values. The blue points, representing the median predictions from 100 bootstrap samples, largely overlap with the original predictions, demonstrating that the model is stable across resampled datasets. The red vertical lines indicate the 89% confidence intervals, which are relatively narrow for most observations, reflecting low prediction uncertainty, but are wider for a few individuals, likely corresponding to less typical combinations of predictor values. Overall, the pattern suggests that the model captures the main trends in the data, and the bootstrap confirms both the stability and the uncertainty of the predictions.

# Final Evaluation

# Predictions for Training Data Using Model 2
```{r}
train_pred_all <- predict(lm_all_fit, new_data = train_data) %>%
  bind_cols(train_data %>% select(Y)) %>%
  mutate(dataset = "Training")
```

# Predictions for Test Data Using Model 2
```{r}
test_pred_all <- predict(lm_all_fit, new_data = test_data) %>%
  bind_cols(test_data %>% select(Y)) %>%
  mutate(dataset = "Test")
```

# Combine Training and Testing Data Predictions
```{r}
final_pred_df <- bind_rows(train_pred_all, test_pred_all)
```

# Plot for Observed vs Predicted for Both Datasets
```{r}
ggplot(final_pred_df, aes(x = Y, y = .pred, color = dataset, shape = dataset)) +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(0, 5000), ylim = c(0, 5000)) +
  labs(
    x = "Observed Values",
    y = "Predicted Values",
    color = "Dataset",
    shape = "Dataset",
    title = "Observed vs Predicted Values for Training and Test Data"
  ) +
  theme_minimal()
```

This plot compares observed vs predicted values for both the training and test data using model 2. The test data points and the training data points are mixed in with each other, which suggests that the model generalizes well to new data and does not overfit.

# Overall Model Assessment

The null model gives us a baseline to compare against, because it does not use any predictors and just predicts the same average value every time. Model 1 uses dose as the only predictor, so the main question is whether dose alone improves prediction. Model 1 has a lower RMSE than the null model, so dose does help explain some of the variation in the outcome. Model 2 includes all predictors, and this model has an even lower RMSE value compared to model 1. The lower RMSE value shows that using all predictors is more accurate. The test data predictions are mixed in with the training data predictions, which means that the model is generalizing to new data rather than overfitting.