---
title: "Fitting Exercise"
---

# Load and summarize data
```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(GGally)
library(tidymodels)

mavo <- read_csv("Mavoglurant.csv")

glimpse(mavo)
dim(mavo)
```

# Convert data to plot
```{r}
ggplot(mavo, aes(x = TIME, y = DV, group = ID, color = factor(DOSE))) +
  geom_line(alpha = 0.6) +
  labs(color = "Dose") +
  theme_minimal()

#only keeping observations with OCC = 1
mavo_occ1 <- mavo %>%
  filter(OCC == 1)

table(mavo_occ1$OCC)
```

# Creates a data frame with time that is not equal to 0 and computes the sum of DV
```{r}
mavo_no0 <- mavo_occ1 %>%
  filter(TIME != 0)

Y_df <- mavo_no0 %>%
  group_by(ID) %>%
  summarize(Y = sum(DV, na.rm = TRUE))
```

# Creates a data frame with time = 0
```{r}
baseline_df <- mavo_occ1 %>%
  filter(TIME == 0)
```

# Joins the tow data frames together
```{r}
final_df <- baseline_df %>%
  left_join(Y_df, by = "ID")

dim(final_df)
```

# Coverts race and sex to variable factors and cleans data
```{r}
clean_df <- final_df %>%
  mutate(
    SEX = factor(SEX),
    RACE = factor(RACE)
  ) %>%
  select(Y, DOSE, AGE, SEX, RACE, WT, HT)

summary(clean_df)
```

# Summary of data
```{r}
clean_df %>%
  summarize(
    n = n(),
    mean_Y = mean(Y),
    sd_Y = sd(Y),
    mean_age = mean(AGE),
    mean_WT = mean(WT),
    mean_HT = mean(HT)
  )
```
This summary table shows that there are 120 indivduals and it shows the the mean and standard deviation of drug exposure, and the mean od age, WT, and HT.

# Summary grouped by dose 
```{r}
  clean_df %>%
  group_by(DOSE) %>%
  summarize(
    n = n(),
    mean_Y = mean(Y),
    sd_Y = sd(Y),
    mean_WT = mean(WT),
    mean_AGE = mean(AGE)
  )
```
This summary table shows the relationship between dose and exposure. As dose increases, total exposure decreases.

# Summary grouped by sex
```{r}
  clean_df %>%
  group_by(SEX) %>%
  summarize(
    n = n(),
    mean_Y = mean(Y),
    mean_WT = mean(WT)
  )
```
This summary table shows the data grouped by sex. Sex 1 has 104 individuals and sex 2 has 16 individuals. 

# Plot y versus dose
```{r}
ggplot(clean_df, aes(x = factor(DOSE), y = Y)) +
  geom_boxplot() +
  labs(x = "Dose", y = "Total Drug (Y)") +
  theme_minimal()
```
This boxplot shows total drug exposure versus dose; the higher the dose is associated with drug exposure.

# Plot y versus age
```{r}
  ggplot(clean_df, aes(x = AGE, y = Y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()
```
This scatterplot shows total drug exposure versus age; there is no relationship between the two variables.

# Distribution of Y
```{r}
ggplot(clean_df, aes(x = Y)) +
  geom_histogram(bins = 20) +
  theme_minimal()
```
This histogram shows total drug exposure skewed to the right. This means that most individuals have exposure values between 1500 and 3500, but a few individuals have high exposure values.

# Distribution of HT
```{r}
  ggplot(clean_df, aes(x = HT)) +
  geom_histogram(bins = 15) +
  theme_minimal()
```
The histogram shoes height with a normal distribution with most idividuals within 1.60 to 1.90 meters.

# Correlation matrix 
```{r}
numeric_df <- clean_df %>%
  select(Y, AGE, WT, HT)

cor(numeric_df)
```
This correlation matrix shows the strength and whether the relationship is positive or negative between total drug exposure, age, height, and weight. +1 menas there is a positive relationship, 0 means there is no relationship, and -1 means there is a negative relationship.

# Correlation matrix plot plot
```{r}
ggpairs(clean_df,
        columns = c("Y", "AGE", "WT", "HT"),
        aes(color = factor(DOSE)))
```
This correlation matrix plot shows the same relationships as the correlation matrix table.

# Model fitting
```{r}
set.seed(123)

data_split <- initial_split(clean_df, prop = 0.8, strata = SEX)

train_data <- training(data_split)
test_data  <- testing(data_split)
```

# Fitting the linear model to the continous outcome of Y using dose as the main predictor
```{r}
lm_spec <- linear_reg() %>%
  set_engine("lm")

lm_dose <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE)

lm_dose_fit <- fit(lm_dose, data = train_data)

pred_dose <- predict(lm_dose_fit, test_data) %>%
  bind_cols(test_data)

  metrics_dose <- pred_dose %>%
  metrics(truth = Y, estimate = .pred)

metrics_dose
```

# Fitting the linear model to the continous outcome of Y using all the predictors
```{r}
lm_all <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE + AGE + SEX + RACE + WT + HT)

lm_all_fit <- fit(lm_all, data = train_data)

pred_all <- predict(lm_all_fit, test_data) %>%
  bind_cols(test_data)

metrics_all <- pred_all %>%
  metrics(truth = Y, estimate = .pred)

metrics_all
```

# Fitting the logistic model to the continous outcome of sex using dose as the main predictor
```{r}
log_spec <- logistic_reg() %>%
  set_engine("glm")

log_dose <- workflow() %>%
  add_model(log_spec) %>%
  add_formula(SEX ~ DOSE)

log_dose_fit <- fit(log_dose, data = train_data)

pred_log_dose <- predict(log_dose_fit, test_data, type = "prob") %>%
  bind_cols(
    predict(log_dose_fit, test_data),
    test_data
  )

pred_log_dose %>%
  metrics(truth = SEX, estimate = .pred_class)

pred_log_dose %>%
  roc_auc(truth = SEX, .pred_1)
```

# Fitting the logistic model to the continous outcome of Y using all the predictors
```{r}
log_all <- workflow() %>%
  add_model(log_spec) %>%
  add_formula(SEX ~ DOSE + AGE + RACE + WT + HT)

log_all_fit <- fit(log_all, data = train_data)

pred_log_all <- predict(log_all_fit, test_data, type = "prob") %>%
  bind_cols(
    predict(log_all_fit, test_data),
    test_data
  )

pred_log_all %>%
  metrics(truth = SEX, estimate = .pred_class)

pred_log_all %>%
  roc_auc(truth = SEX, .pred_1)
```

The first linear model is using dose as the only predictor of total drug exposure. It shows that the higher the dose, the higher the drug exposure is. 

The second linear model is using all variables (age, dose, sex, race, height, and weight) as predictors of total drug exposure. It showed that dose was the strongest predictor of total drug exposure. 

The first logistic model is using dose as the only predictor of sex. This model produced low ROC-AUC and accuracy values because there is no relationship between dose and sex.

The second logistic model is using all the variables as predictors of sex. This model has higher ROC-AUC and accuracy values compared to the first model becuase height and weight can potentially be predictors of sex.

# Model Improvement 
Excluding race as a variable
```{r}
rngseed = 1234

clean_df <- final_df %>%
  mutate(
    SEX = factor(SEX)
  ) %>%
  select(Y, DOSE, AGE, SEX, WT, HT)

dim(clean_df)
summary(clean_df)
```

Setting test and train data
```{r}
set.seed(rngseed)

data_split <- initial_split(clean_df, prop = 0.75)

train_data <- training(data_split)
test_data  <- testing(data_split)

nrow(train_data)
nrow(test_data)
```

# Linear models
```{r}
lm_spec <- linear_reg() %>%
  set_engine("lm")
```

# Linear Model with Dose
```{r}
lm_dose_workflow <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE)

lm_dose_fit <- fit(lm_dose_workflow, data = train_data)

lm_dose_pred <- predict(lm_dose_fit, new_data = train_data) %>%
  bind_cols(train_data %>% select(Y))

metrics (lm_dose_pred, truth = Y, estimate = .pred)
```

# Linear Model with all Predictors
```{r}
lm_all_workflow <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE + AGE + SEX + WT + HT)

lm_all_fit <- fit(lm_all_workflow, data = train_data)

lm_all_pred <- predict(lm_all_fit, new_data = train_data) %>%
  bind_cols(train_data %>% select(Y))

metrics (lm_all_pred, truth = Y, estimate = .pred)
```

# Null model 
```{r}
mean_y <- mean(train_data$Y)

null_pred <- train_data %>%
  mutate(.pred = mean_y)

rmse_null <- null_pred %>%
  rmse(truth = Y, estimate = .pred)

rmse_null
```

The null model, predicts the mean outcome for all observations, had the highest RSME being 948. This RSME value indicates that the predictive preformace of this model is poor. The model that used dose as the only predictor had a lower RSME value (703), this suggests that dose explains variability in the outcome. The model that uses all predictors had the lowest RSME (627), which indicates the best preformance. 

```{r}
set.seed(rngseed)

# creates 10 fold CV split
folds <- vfold_cv(train_data, v = 10)
```

# Model using Dose
```{r}
lm_dose_workflow <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE)
```

# Model Using All Predictors 
```{r}
lm_all_workflow <- workflow() %>%
  add_model(lm_spec) %>%
  add_formula(Y ~ DOSE + AGE + SEX + WT + HT)
```

# Cross Validation
```{r}
  dose_cv <- fit_resamples(
  lm_dose_workflow,
  resamples = folds,
  metrics = metric_set(rmse)
)

all_cv <- fit_resamples(
  lm_all_workflow,
  resamples = folds,
  metrics = metric_set(rmse)
)
```

# RSME and Standard Validation 
```{r}
dose_cv_results <- collect_metrics(dose_cv)
all_cv_results  <- collect_metrics(all_cv)

dose_cv_results
all_cv_results
```

The RMSE values from the cros validation are a little higher than those from the training data, this could be because of less overfitting. This gives us a more realistic idea of the models preformance on new data. The model that used all the predictors still had the lowest RSME value, followed by the model that used dose as the only predictor, and the null model had the highest RSME value. The standard error values suggest that there is some variability in how the model preforms across folds, but the differences are consistent. 

# Different Value for the Random Seed
```{r}
set.seed(999)

folds2 <- vfold_cv(train_data, v = 10)

dose_cv2 <- fit_resamples(lm_dose_workflow, resamples = folds2, metrics = metric_set(rmse))
all_cv2  <- fit_resamples(lm_all_workflow, resamples = folds2, metrics = metric_set(rmse))

collect_metrics(dose_cv2)
collect_metrics(all_cv2)
```

Changing the random seed resulted in slightly different RSME values, this could be due to variation in how the data was split into the folds. The pattern stayed the same with the model that used all the predictors having the lowest RSME value, and the null model having the highest RSME value. 
