---
title: "Data Analysis"
---
```{r}
#load dslabs package
library(dslabs)
library(tidyverse)

#help(gapminder)
#The str functions is used to look at observations and variables in the dataset  
str(gapminder)

#get a summary of data
summary(gapminder)

#determine the type of object gapminder is
#The class function is a data frame that allows us to look at a class of the dataset
class(gapminder)

#Using the gapminder dataset, I've created a new dataframe that will only include countries that have Africa as the continent
africadata <- gapminder[gapminder$continent == "Africa", ]

#Check the structure and summary of africadata
str(africadata)
summary(africadata)

#Create a dataframe with infant mortality and life expectancy
africa_mort_life <- africadata[, c("infant_mortality", "life_expectancy")]

#Inspect the structure and summary
str(africa_mort_life)
summary(africa_mort_life)

#Create a dataframe with population and life expectancy
africa_pop_life <- africadata[, c("population", "life_expectancy")]

#Inspect the structure and summary
str(africa_pop_life)
summary(africa_pop_life)
```

This analysis looks at African countries using the Gapminder dataset on R. The data is first subset to only include African countries. Then two smaller datasets are created to look at life expectancy with infant mortality and population with infant mortality. 

```{r}
#Plot life expectancy vs infant mortality for African countries
#Using the plot function to make the plot
plot(
    #Using the infant_mortality column and makes them the x-axis values
  africa_mort_life$infant_mortality,
  #Using the life_expectancy column and makes them the y-axis values
  africa_mort_life$life_expectancy,
  #labels the x-axis infant mortality and the y-axis Life expectancy
  xlab = "Infant Mortality",
  ylab = "Life Expectancy",
  #Makes the title of the plot Life expectancy vs infant mortality in Africa
  main = "Life Expectancy vs Infant Mortality in Africa",
  pch = 16
)


#Plot life expectancy vs population size for African countries
#Using the plot function to make the plot
plot(
    #Using the population column and makes them the x-axis values
  africa_pop_life$population,
  #Using the life expectancy column and makes them the y-axis values
  africa_pop_life$life_expectancy,
  #labels the x-axis population size and the y-axis Life expectancy
  xlab = "Population Size (log scale)",
  ylab = "Life Expectancy",
  #Makes the title of the plot Life expectancy vs population size in Africa
  main = "Life Expectancy vs Population Size in Africa",
  pch = 16,
  #Setting the scale of the x-axis to be logarithmic
  log = "x"
)
```

The first plot shows a negative relationship between infant mortality and life expectancy. The second plot shows a positive relationship between population size and life expectancy. The streaks in the plots are most likley due to individual countries obsereved over multiple years. This shows more in the second plot because there is data for the same countries over several years, which is likely to show gradual differences. These gradual differences may form streaks in the plots.

```{r}
#Identifies which years have missing data for infant mortality 
#is.na() functions chacks if each value in the dataframe is NA or not
missing_im <- is.na(africadata$infant_mortality)

#Lists the years when infant mortality is missing
missing_years <- sort(unique(africadata$year[missing_im]))
missing_years

#Extract data for the year 2000 only
africa_2000 <- africadata[africadata$year == 2000, ]

#Check the structure of the new dataset
str(africa_2000)

#View summary statistics for the year 2000 data
summary(africa_2000)

#Plot life expectancy vs infant mortality for Africa in the year 2000
#Using the plot function to make the plot
plot(
     #Using the infant mortality column and makes them the x-axis values
  africa_2000$infant_mortality,
   #Using the life expectancy column and makes them the y-axis values
  africa_2000$life_expectancy,
  #labels the x-axis infant mortality and the y-axis Life expectancy
  xlab = "Infant Mortality",
  ylab = "Life Expectancy",
  #Makes the title of the plot Life expectancy vs infant mortality in Africa in 2000
  main = "Life Expectancy vs Infant Mortality in Africa in 2000",
  pch = 16
)

# Plot life expectancy vs population size for Africa in the year 2000
#Using the plot function to make the plot
plot(
     #Using the population column and makes them the x-axis values
  africa_2000$population,
   #Using the life expectancy column and makes them the y-axis values
  africa_2000$life_expectancy,
  #labels the x-axis population size and the y-axis Life expectancy
  xlab = "Population Size (log scale)",
  ylab = "Life Expectancy",
  #Makes the title of the plot Life expectancy vs population size in Africa in 2000
  main = "Life Expectancy vs Population Size in Africa in 2000",
  pch = 16,
  #Setting the scale of the x-axis to be logarithmic
  log = "x"
)

#Models life expectancy as a function of infant mortality
#lm is used to fit linear models
lm_infant <- lm(life_expectancy ~ infant_mortality, data = africa_2000)

# Models life expectancy as a function of population size
lm_population <- lm(life_expectancy ~ population, data = africa_2000)

# prints results for both models
summary(lm_infant)
summary(lm_population)
```

The relationship between life expectancy and infant mortality are highly significant, and negatively correlated. The negative correlation is indicated by the negative slope. The relationship between population size and life expectancy is not significant because the p-value is 0.6159.